Supplementary Information: Tropical tree drought sensitivity is jointly shaped by drought characteristics, species adaptations, and individual microenvironments

### 2025-04-10 doc/manuscript_hkk_drought_sensitivity.Rmd
Update manuscript_hkk_drought_sensitivity.Rmd
@krishnaanujan - I edited the latter part of methods, results, and 1st par of discussion.

<!--PARAGRAPH ON HOW MICROENVIRONMENT SHAPES DROUGHT RESPONSES-->
Finally, the microenvironments that individual trees experience influence growth responses to drought. <!--leading to varied atmospheric and hydrological stresses with distinct ecological consequences.-->
Across horizontal axes, microenvironments defined by topographic and edaphic variation shape tree drought responses. 
While habitats with relatively moist soils (e.g., low-lying topographic positions) are clearly more hospitable during drought, trees in these microenvironments may be more vulnerable if less drought-adapted [e.g., less negative $\pi_{tlp}$, @kunert_leaf_2021] or acclimatized [@refs].
Indeed, there are observations of trees near streams undergoing greater growth declines [@mcgregor_tree_2021] and increases in mortality [@zuleta_droughtinduced_2017] during drought. 
Across vertical axes, trees of different sizes have their crowns and root systems in sometimes dramatically divergent microenvironments.
Aboveground, tree crowns in canopy or emergent positions are exposed to higher evaporative demand and thermal stress driven by higher solar radiation, wind speed, and vapor pressure deficit [@vinod_thermal_2023], and such exposure makes trees more vulnerable to drought [@scharnweber_confessions_2019; @ma_tree_2023].
<!--In contrast, understory trees may be rarely exposed to direct solar radiation and extreme evaporative demand.-->
These aboveground gradients in crown exposure are at least partially offset by the tendency for large trees to have larger and often deeper root systems [@refs] and thereby potential to access more reliable water sources [@refs].
However, trees with deeper roots that rely on regular access to deep water may be more vulnerable during severe droughts when those sources are depleted [@chitra-tarak_hydraulicallyvulnerable_2021]. 
While the overall tendency is that larger trees suffer greater growth declines during drought [@bennett_larger_2015], it remains unclear whether this is driven primarily by crown exposure, by greater height itself makes trees more vulnerable to drought based on the physics of hydraulic flow through a porous medium<!--, as described by Darcy's law -->[@mcdowell_darcy_2015; @fernández-de-uña_role_inpress], or by trait differences across vertical axis [@vinod_thermal_2023].
Thus, horizontal and vertical microenvironmental gradients modify tree drought responses, likely in interaction with drought characteristics and species traits.

@@

![***Figure `r fig_droughts`. Climatic characteristics of drought years and average years*** in the Huai Kha Khaeng ForestGEO plot. For four daily climate variables dry days, precipitation, maximum temperature and VPD, panel a) represents rolling means for 30 days around the focal day for that year. Grey line represents the mean of this rolling mean from 2008 to 2019 and ribbons represent standard deviation around this mean. Panel b) represents anomalies of this rolling mean from the values across the timeseries. Values, therefore represents the number of standard deviations from the mean on a given day. c) shows Standardised Precipitation Evapotranspiration Index (SPEI) values for each month in 2010 and 2015. Across all panels, light blue background represents the average wet season, from May to October. For comments - https://github.com/forestgeo/growth-precip-thailand/issues/10](display/Fig1.png)

@@

![***Figure `r fig_drought_responses`: Growth increment timeseries and drought sensitivity*** across trees and species in the Huai Kha Khaeng ForestGEO plot. a) Median of annualised growth increments across species (colours) and the whole plot (black) from 2009 - 2022 from late wet/early dry season dendroband measurements. 2010 and 2015 were moderate and very strong ENSO events respectively. b) Distribution of drought sensitivity of tree growth across all individuals in the two drought years. For comments - https://github.com/forestgeo/growth-precip-thailand/issues/11](display/Fig2.png)

@@

We identified drought years of interest in the dendroband timeseries - 2010 and 2015 - based on two drought indices and corroborated by meteorological data, local expertise, and plot-level growth responses. 
Within the dendroband timeseries, two ENSO events had occurred; 2010 was a moderate ENSO event while 2015 was a very strong ENSO event [@REF]. 
<!-- We chose to identify the years starting from the timeseries, instead of directly using climate variables and their deviation because drought definitions vary widely, and drought characterisation of a year can vary based on the definition selected. -->
We also calculate Standardised Precipitation-Evapotranspiration Index (SPEI) at the monthly scale using the SPEIbase dataset [@vicente-serrano_multiscalar_2010].
Defining drought as periods with two or more consecutive months with SPEI < -1 also resulted in identification of 2010 and 2015 as drought events. 
These years also corresponded with expert knowledge of major droughts that affected the region as well as meteorological records (see next paragraph; Fig. `r fig_droughts`) and plot-level growth declines in the timeseries (Fig. `r fig_drought_responses`). 
<!-- We report monthly SPEI for the two drought years along with long-term mean SPEI for each month.  -->
<!-- Moreover, broad definitions based on climate variables (often not measured at the location), may not capture ecological realities within the plot.  -->
We note that although tree growth was unusally low in 2020, this year did not emerge as a drought year either in the examination of drought indicators or from expert knowledge.

@@

To account for potential biases in dendrometer band increments from water-related growth or shrinkage, we also calculated growth occurrence using a zero-growth assumption [@zweifel_are_2016]. 
Timing of measurement and seasonal dynamics in tree hydration and bark dynamics could contribute to negative values in dendrometer band increments, unlike woody growth from tree ring datasets [@chitra-tarak_it_2015]. 
Using a zero-growth model, we calculated growth occurrence as 1 under positive diameter change and any change <=0 as 0. 
To account for biases introduced by timing of measurements and seasonal dynamics, we used this simplifying assumption to corroborate the patterns observed with the sensitivity metric.

@@

We calculated deciduousness as a species level trait using data published in @williams_deciduousness_2008 for the top 30 species. 
We defined deciduousness as the mean proportion of canopy loss at maximum loss across individuals in the species, with a score ranging from 1 to 4 where 1 is most evergreen and 4 is most deciduous.  
Data was unavailable for _Alphonsea ventricosa_, an evergreen species, and so we manually assigned it a deciduousness score of 1, in the middle of the evergreen range. 
An alternate measure of deciduousness is the duration spent at maximum crown loss.
This data was strongly correlated with the proportion of crown loss; we chose to use the proportion of crown loss because it was available for a larger list of species.  
<!-- At the species level, we also calculated the maximum size across all individuals in the plot as measured in the 2008 whole-plot census.
We calculated niche breadth as the standard deviation of TWI across all observed individuals of each species in the whole plot census in 2008.  -->

@@

The two droughts analysed had distinct temporal climatic characteristics, with 2010 primarily a dry season drought and 2015 a wet-season drought (Table 1, Fig `r fig_droughts`).
In 2010, drought characteristics of lower precipitation, higher number of dry days, higher maximum temperature, higher VPD than the long-term mean and SPEI values below -1 occurred between February and April, in the dry season. 
On the other hand, in 2015, similar deviations occurred in May and June, in the wet season. 
In 2015, SPEI in May was -2, an anomalously low value that was not observed in any month in the entire dendroband census time period, suggesting a severe drought.

@@

## Growth timeseries and Drought sensitivity {-}

The increment timeseries using dendrometer band measurements revealed plot- and species-level patterns in interannual growth variation (Fig `r fig_drought_responses`a). 
Median annualised growth rate across the years of measurement for the whole plot was `r round(median(tree.time$inc_annual, na.rm=T),2)` $\pm$ `r round(sd(tree.time$inc_annual, na.rm=T),2)` cm. 
Median annual increment differed by species with *`r tree.time.sp$spfull[1]`* having the lowest (`r tree.time.sp$inc_annual_med[1]` $\pm$ `r tree.time.sp$inc_annual_sd[1]` cm) and *`r tree.time.sp$spfull[nrow(tree.time.sp)]`* the highest (`r tree.time.sp$inc_annual_med[nrow(tree.time.sp)]` $\pm$ `r tree.time.sp$inc_annual_sd[nrow(tree.time.sp)]` cm) among the `r nrow(tree.time.sp)` species analysed.

Drought years were associated with lower growth across the plot. 
The drought year 2015 had the lowest median annual increment `r round(median(tree.time$inc_annual[tree.time$yr==2015], na.rm=T),2)` cm, `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2015], na.rm=T)-median(tree.time$inc_annual, na.rm=T))/median(tree.time$inc_annual, na.rm=T)),2)`% lower than the median across all years. 
In 2010, the median annual increment was `r round(median(tree.time$inc_annual[tree.time$yr==2010], na.rm=T),2)` cm, `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2010], na.rm=T)-median(tree.time$inc_annual, na.rm=T))/median(tree.time$inc_annual, na.rm=T)),2)`% lower than the median across all years. 
While there were more species differences in median responses to the 2010 drought, all species had lower increment than a typical year in 2015. 
Besides decrease in magnitude of growth, the proportion of trees with negative growth was also higher in the drought years (Fig S4). 
Across the years, on average, `r round(100*mean(tapply(tree.time$inc_annual, tree.time$yr, function(x) sum(x < 0) / length(x))),2)`% of the trees had negative growth, with a minimum of `r round(100*min(tapply(tree.time$inc_annual, tree.time$yr, function(x) sum(x < 0) / length(x))),2)`% in `r names(sort(tapply(tree.time$inc_annual, tree.time$yr, function(x) sum(x < 0) / length(x))))[1]`. The drought year 2015 had the maximum proportion of trees with negative growth in the timeseries, `r round(100*sum(tree.time$inc_annual[tree.time$yr==2015]<0)/nrow(tree.time[tree.time$yr==2015,]),2)`%.

Trees on average had negative drought sensitivities for both droughts with considerable variation around it (Fig `r fig_drought_responses`b). 
Median drought sensitivity across all trees in 2010 was `r round(median(tree.time$sens.prop[tree.time$yr==2010]), 2)` $\pm$ `r round(sd(tree.time$sens.prop[tree.time$yr==2010]), 2)` while in 2015 it was `r round(median(tree.time$sens.prop[tree.time$yr==2015]), 2)` $\pm$ `r round(sd(tree.time$sens.prop[tree.time$yr==2015]), 2)`. 
In 2010, *`r sens.sp[sens.sp$yr==2010,]$spfull[1]`* and *`r sens.sp[sens.sp$yr==2010,]$spfull[nrow(sens.sp)/2]`* had the lowest and highest median sensitivity of `r sens.sp[sens.sp$yr==2010,]$sens_med[1]` and `r sens.sp[sens.sp$yr==2010,]$sens_med[nrow(sens.sp)/2]` respectively. 
In 2015, *`r sens.sp[sens.sp$yr==2015,]$spfull[1]`* and *`r sens.sp[sens.sp$yr==2015,]$spfull[nrow(sens.sp)/2]`* had the lowest and highest median sensitivity of `r sens.sp[sens.sp$yr==2015,]$sens_med[1]` and `r sens.sp[sens.sp$yr==2015,]$sens_med[nrow(sens.sp)/2]` respectively.

@@

## Species effects on drought sensitivity {-}

Deciduousness was associated with lower negative sensitivities in 2015, but not in 2010 (Fig `r fig_species`a). 
Species median sensitivities in 2010 and 2015 were not correlated with each other (Pearson's correlation coefficient r = `r round(sens.sp.cor[4]$estimate, 2)`, p = `r round(sens.sp.cor[3]$p.value,2)`, Fig S5) but on average had lower mean and median sensitivities in 2015 than in 2010 (Fig S5), in alignment with median sensitivities across all trees. 
In 2015, predicted sensitivities from intercept-only models of sensitivity with a species random effect on the intercept showed a significant negative association with deciduousness value (r = `r round(coefs_dec_lms$estimate[coefs_dec_lms$yr==2015], 2)`, p = `r round(coefs_dec_lms$p.value[coefs_dec_lms$yr==2015], 2)`). 
On the other hand, in 2010, predicted sensitivities had no significant association with deciduousness (r = `r round(coefs_dec_lms$estimate[coefs_dec_lms$yr==2010], 2)`, p = `r round(coefs_dec_lms$p.value[coefs_dec_lms$yr==2010], 2)`). 
Moreover, models of all trees with TWI and deciduousness as predictors showed that the conditional effect of deciduousness on sensitivity (controlling for topographic wetness) was significant and negative in 2015 (`r round(isocline_coefs_dec$median[isocline_coefs_dec$yr==2015],2)`; 90% CI `r round(isocline_coefs_dec$lwr[isocline_coefs_dec$yr==2015],2)`, `r round(isocline_coefs_dec$upr[isocline_coefs_dec$yr==2015],2)`) but not in 2010 (`r round(isocline_coefs_dec$median[isocline_coefs_dec$yr==2010],2)`, 90% CI `r round(isocline_coefs_dec$lwr[isocline_coefs_dec$yr==2010],2)`, `r round(isocline_coefs_dec$upr[isocline_coefs_dec$yr==2010],2)`).

## Microenvironment effects drought sensitivity {-}

The effect of wetness on sensitivity was different in 2010 and 2015, with 2015 effects interacting with deciduousness. 
Individual tree responses to drought were not correlated with each other (Fig S6). 
Across all trees, the model with TWI and deciduousness showed that TWI had no effect on drought sensitivity in 2010 (median effect =`r round(isocline_coefs_twi$median[isocline_coefs_twi$yr==2010],2)`, 90% CI `r round(isocline_coefs_twi$lwr[isocline_coefs_twi$yr==2010],2)`, `r round(isocline_coefs_twi$upr[isocline_coefs_twi$yr==2010],2)`) but a positive effect in 2015 (median effect =`r round(isocline_coefs_twi$median[isocline_coefs_twi$yr==2015],2)`, 90% CI `r round(isocline_coefs_twi$lwr[isocline_coefs_twi$yr==2015],2)`, `r round(isocline_coefs_twi$upr[isocline_coefs_twi$yr==2015],2)`). 
The interaction between TWI and deciduousness was not significant in 2010 (median effect =`r round(isocline_coefs_inter$median[isocline_coefs_inter$yr==2010],2)`, 90% CI `r round(isocline_coefs_inter$lwr[isocline_coefs_inter$yr==2010],2)`, `r round(isocline_coefs_inter$upr[isocline_coefs_inter$yr==2010],2)`) but was significantly negative in 2015 (median effect =`r round(isocline_coefs_inter$median[isocline_coefs_inter$yr==2015],2)`, 90% CI `r round(isocline_coefs_inter$lwr[isocline_coefs_inter$yr==2015],2)`, `r round(isocline_coefs_inter$upr[isocline_coefs_inter$yr==2015],2)`).
Combined, predicted sensitivities of deciduous trees was more negative than evergreen trees in 2015, with this effect stronger in wetter sites (Fig `r fig_species`b).

The opposite effects of TWI on sensitivity in the wet and dry season drought, and the influence of deciduousness was also confirmed by models that controlled for crown exposure and size as well as species variation in intercept and slopes (Fig `r fig_microenvs`a). 
These models showed small negative sensitivities in wetter sites in the dry season drought (median effect =`r round(coefs_spre_twi$median[coefs_spre_twi$yr==2010],2)`, 90% CI `r round(coefs_spre_twi$lwr[coefs_spre_twi$yr==2010],2)`, `r round(coefs_spre_twi$upr[coefs_spre_twi$yr==2010],2)`), but strong positive effect of wetness in the wet season drought (median effect =`r round(coefs_spre_twi$median[coefs_spre_twi$yr==2015],2)`, 90% CI `r round(coefs_spre_twi$lwr[coefs_spre_twi$yr==2015],2)`, `r round(coefs_spre_twi$upr[coefs_spre_twi$yr==2015],2)`). 
The positive effect of TWI in 2015 was, however, mainly due to the response of evergreen species; deciduousness had a significant negative correlation with the slopes of the TWI effect in 2015 (Fig `r fig_microenvs`b). 
Predictions across the range of observed variables also showed that deciduous species generally had larger negative sensitivities in 2015, which did not show increase with wetness (Fig `r fig_microenvs`c). 
On the other hand, neither the mean sensitivities across species nor their response to TWI was associated with deciduousness in 2010.

![***Figure `r fig_microenvs`: Drivers of variation in sensitivity among individuals***. All panels represent model results from models (one for each year) predicting sensitivity with microenvironment conditions with species random effect on all slopes. a) Coefficient plots showing median effects and 90% CI for wetness, exposure and DBH of tree. b) wetness effect across all species and its relationship with deciduousness value. Line represents a significant correlation. c) Predicted relationship of drought sensitivity with topographic wetness index across all species derived from model predictions. Black line represents overall relationship; coloured lines for each species reflects deciduousness values. For comments - https://github.com/forestgeo/growth-precip-thailand/issues/13](display/Fig4.png)

@@

Accounting for effects of exposure, remaining effects of size on sensitivity differed between the two droughts. 
DBH was a significant predictor of exposure category, with larger DBH associated with higher exposure categories in 2010 (median effect =`r round(coefs_spre_ordcii$median[coefs_spre_ordcii$yr==2010],2)`, 90% CI `r round(coefs_spre_ordcii$lwr[coefs_spre_ordcii$yr==2010],2)`, `r round(coefs_spre_ordcii$upr[coefs_spre_ordcii$yr==2010],2)`) and 2015 (median effect =`r round(coefs_spre_ordcii$median[coefs_spre_ordcii$yr==2015],2)`, 90% CI `r round(coefs_spre_ordcii$lwr[coefs_spre_ordcii$yr==2015],2)`, `r round(coefs_spre_ordcii$upr[coefs_spre_ordcii$yr==2015],2)`) (Fig 5a). 
Controlling for this direct effect of exposure, the remaining effect of size *per se* was significantly positive in 2010, while it was weakly negative in 2015 (Fig `r fig_microenvs`a, Fig `r fig_dag`a).

>>>>

<!--PARAGRAPH ON HOW MICROENVIRONMENT SHAPES DROUGHT RESPONSES-->
Finally, the microenvironments that individual trees experience influence growth responses to drought. <!--leading to varied atmospheric and hydrological stresses with distinct ecological consequences.-->
Across horizontal axes, microenvironments defined by topographic and edaphic variation shape tree drought responses. 
While habitats with relatively moist soils (e.g., low-lying topographic positions) are clearly more hospitable during drought, trees in these microenvironments may be more vulnerable if less drought-adapted [e.g., less negative $\pi_{tlp}$, @kunert_leaf_2021] or acclimatized [@refs].
Indeed, there are observations of trees near streams undergoing greater growth declines [@mcgregor_tree_2021] and increases in mortality [@zuleta_droughtinduced_2017] during drought. 
Across vertical axes, trees of different sizes have their crowns and root systems in sometimes dramatically divergent microenvironments.
Aboveground, tree crowns in canopy or emergent positions are exposed to higher evaporative demand and thermal stress driven by higher solar radiation, wind speed, and vapor pressure deficit [@vinod_thermal_2023], and such exposure makes trees more vulnerable to drought [@scharnweber_confessions_2019; @ma_tree_2023].
<!--In contrast, understory trees may be rarely exposed to direct solar radiation and extreme evaporative demand.-->
These aboveground gradients in crown exposure are at least partially offset by the tendency for large trees to have larger and often deeper root systems [@refs] and thereby potential to access more reliable water sources [@refs].
However, trees with deeper roots that rely on regular access to deep water may be more vulnerable during severe droughts when those sources are depleted [@chitra-tarak_hydraulicallyvulnerable_2021]. 
While the overall tendency is that larger trees suffer greater growth declines during drought [@bennett_larger_2015], it remains unclear whether this is driven primarily by crown exposure, by greater height itself makes trees more vulnerable to drought based on the physics of hydraulic flow through a porous medium <!--, as described by Darcy's law -->[@mcdowell_darcy_2015; @fernández-de-uña_role_inpress], or by trait differences across vertical axis [@vinod_thermal_2023].
Thus, horizontal and vertical microenvironmental gradients modify tree drought responses, likely in interaction with drought characteristics and species traits.

@@

![***Figure `r fig_droughts`. Climatic characteristics of drought years and average years*** in the Huai Kha Khaeng ForestGEO plot. For four daily climate variables dry days, precipitation, maximum temperature and VPD, panel a) represents rolling means for 30 days around the focal day for that year. Grey line represents the mean of this rolling mean from 2008 to 2019 and ribbons represent standard deviation around this mean. Panel b) represents anomalies of this rolling mean from the values across the timeseries. Values, therefore represents the number of standard deviations from the mean on a given day. c) shows 1-month Standardised Precipitation Evapotranspiration Index (SPEI) values for each month in 2010 and 2015. Across all panels, light blue background represents the average wet season, from May to October. For comments - https://github.com/forestgeo/growth-precip-thailand/issues/10](display/Fig1.png)

@@

![***Figure `r fig_drought_responses`: Growth increment timeseries and drought responses*** across trees and species in the Huai Kha Khaeng ForestGEO plot. a) Median of annualised growth increments across ten dominant species (colours) and all trees (black) from 2009 - 2022. b) Distribution of drought sensitivity of tree growth across all individuals in the two drought years. For comments - https://github.com/forestgeo/growth-precip-thailand/issues/11](display/Fig2.png)

@@

We identified drought years of interest in the dendroband timeseries - 2010 and 2015 - based on two drought indices and corroborated by meteorological data, local expertise, and plot-level growth responses. 
Within the dendroband timeseries, two ENSO events had occurred; 2010 was a moderate ENSO event while 2015 was a very strong ENSO event [@REF]. 
<!-- We chose to identify the years starting from the timeseries, instead of directly using climate variables and their deviation because drought definitions vary widely, and drought characterisation of a year can vary based on the definition selected. -->
We also calculate Standardised Precipitation-Evapotranspiration Index (SPEI) at the 1-month scale using the SPEIbase dataset [@vicente-serrano_multiscalar_2010].
Defining drought as periods with two or more consecutive months with 1-month SPEI < -1 also resulted in identification of 2010 and 2015 as drought events. 
These years also corresponded with expert knowledge of major droughts that affected the region as well as meteorological records (see next paragraph; Fig. `r fig_droughts`) and plot-level growth declines in the timeseries (Fig. `r fig_drought_responses`). 
<!-- We report monthly SPEI for the two drought years along with long-term mean SPEI for each month.  -->
<!-- Moreover, broad definitions based on climate variables (often not measured at the location), may not capture ecological realities within the plot.  -->
We note that although tree growth was unusally low in 2020, this year did not emerge as a drought year either in the examination of drought indicators or from expert knowledge.

@@

*(See [issue #28](https://github.com/forestgeo/growth-precip-thailand/issues/28))*
To account for potential biases in dendrometer band increments from water-related growth or shrinkage, we also calculated growth occurrence using a zero-growth assumption [@zweifel_are_2016]. 
Timing of measurement and seasonal dynamics in tree hydration and bark dynamics could contribute to negative values in dendrometer band increments, unlike woody growth from tree ring datasets [@chitra-tarak_it_2015]. 
Using a zero-growth model, we calculated growth occurrence as 1 under positive diameter change and any change <=0 as 0. 
To account for biases introduced by timing of measurements and seasonal dynamics, we used this simplifying assumption to corroborate the patterns observed with the sensitivity metric.

@@

We calculated deciduousness as a species level trait using data published in @williams_deciduousness_2008 for the top 30 species. 
We defined deciduousness as the mean proportion of crown leaf loss at maximum loss across individuals in the species, with a score ranging from 1 to 4 where 1 is most evergreen and 4 is most deciduous.
Data was unavailable for _Alphonsea ventricosa_, an evergreen species, and so we manually assigned it a deciduousness score of 1, in the middle of the evergreen range. 
An alternate measure of deciduousness is the duration spent at maximum crown loss.
This was strongly correlated with the proportion of crown loss; we chose to use the proportion of crown loss because it was available for a larger list of species.  
<!-- At the species level, we also calculated the maximum size across all individuals in the plot as measured in the 2008 whole-plot census.
We calculated niche breadth as the standard deviation of TWI across all observed individuals of each species in the whole plot census in 2008.  -->

@@

The two droughts analysed had distinct temporal climatic characteristics, with 2010 primarily a dry season drought and 2015 an early wet-season drought (Table 1, Fig `r fig_droughts`).
The 2010 drought peaked between February and April (dry season), with drought characteristics including lower precipitation, higher number of dry days, higher maximum temperature, and higher VPD relative to the long-term mean and 1-month SPEI values below -1. 
In 2015, similar deviations occurred in May and June (early wet season), with 1-month SPEI in May reaching -2, the lowest in the the entire dendroband census time period.

@@

## Growth timeseries and drought responses {-}

The increment timeseries using dendrometer band measurements revealed plot- and species-level patterns in interannual growth variation (Fig `r fig_drought_responses`a). 
Median annualised growth rate across the years of measurement for the whole plot was `r round(median(tree.time$inc_annual, na.rm=T),2)` $\pm$ `r round(sd(tree.time$inc_annual, na.rm=T),2)` cm. 
Median annual increment differed by species with *`r tree.time.sp$spfull[1]`* having the lowest (`r tree.time.sp$inc_annual_med[1]` $\pm$ `r tree.time.sp$inc_annual_sd[1]` cm) and *`r tree.time.sp$spfull[nrow(tree.time.sp)]`* the highest (`r tree.time.sp$inc_annual_med[nrow(tree.time.sp)]` $\pm$ `r tree.time.sp$inc_annual_sd[nrow(tree.time.sp)]` cm) among the `r nrow(tree.time.sp)` species analysed. 
*(See [issue #27](https://github.com/forestgeo/growth-precip-thailand/issues/27))*

Drought years were associated with lower growth across the plot. 
In 2010, the median annual increment was `r round(median(tree.time$inc_annual[tree.time$yr==2010], na.rm=T),2)` cm, `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2010], na.rm=T)-median(tree.time$inc_annual, na.rm=T))/median(tree.time$inc_annual, na.rm=T)),2)`% lower than the median across all years. 
The drought year 2015 had the lowest median annual increment `r round(median(tree.time$inc_annual[tree.time$yr==2015], na.rm=T),2)` cm, `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2015], na.rm=T)-median(tree.time$inc_annual, na.rm=T))/median(tree.time$inc_annual, na.rm=T)),2)`% lower than the median across all years. 
While median species responses to the 2010 drought were variable, all species had lower increment than a typical year in 2015. 
Besides decrease in magnitude of growth, the proportion of trees with negative growth was also higher in the drought years (Fig S4). 
Across the years, on average, `r round(100*mean(tapply(tree.time$inc_annual, tree.time$yr, function(x) sum(x < 0) / length(x))),2)`% of the trees had negative growth, with a minimum of `r round(100*min(tapply(tree.time$inc_annual, tree.time$yr, function(x) sum(x < 0) / length(x))),2)`% in `r names(sort(tapply(tree.time$inc_annual, tree.time$yr, function(x) sum(x < 0) / length(x))))[1]`. 
The drought year 2015 had the maximum proportion of trees with negative growth in the timeseries, `r round(100*sum(tree.time$inc_annual[tree.time$yr==2015]<0)/nrow(tree.time[tree.time$yr==2015,]),2)`%.
*(See [issue #28](https://github.com/forestgeo/growth-precip-thailand/issues/28))*

On average, trees had negative drought sensitivities for both droughts, but with considerable variation (Fig `r fig_drought_responses`b). 
Median drought sensitivity across all trees in 2010 was `r round(median(tree.time$sens.prop[tree.time$yr==2010]), 2)` $\pm$ `r round(sd(tree.time$sens.prop[tree.time$yr==2010]), 2)` while in 2015 it was `r round(median(tree.time$sens.prop[tree.time$yr==2015]), 2)` $\pm$ `r round(sd(tree.time$sens.prop[tree.time$yr==2015]), 2)`. 
In 2010, species-level sensitivities ranged from `r sens.sp[sens.sp$yr==2010,]$sens_med[1]` for *`r sens.sp[sens.sp$yr==2010,]$spfull[1]`* to `r sens.sp[sens.sp$yr==2010,]$sens_med[nrow(sens.sp)/2]` for *`r sens.sp[sens.sp$yr==2010,]$spfull[nrow(sens.sp)/2]`*. 
In 2015, these ranged from `r sens.sp[sens.sp$yr==2015,]$sens_med[1]` for *`r sens.sp[sens.sp$yr==2015,]$spfull[1]`* to `r sens.sp[sens.sp$yr==2015,]$sens_med[nrow(sens.sp)/2]` for *`r sens.sp[sens.sp$yr==2015,]$spfull[nrow(sens.sp)/2]`*.

@@

## Species effects {-}

Species median sensitivities in 2010 and 2015 were not correlated with each other (Pearson's correlation coefficient r = `r round(sens.sp.cor[4]$estimate, 2)`, p = `r round(sens.sp.cor[3]$p.value,2)`, Fig S5) but on average had lower mean and median sensitivities in 2015 than in 2010 (Fig S5), in alignment with median sensitivities across all trees.

Deciduousness species had lower negative sensitivities in 2015, but not in 2010 (Fig `r fig_species`a). 
In 2015, predicted sensitivities from intercept-only models of sensitivity with a species random effect on the intercept showed a significant negative association with deciduousness value (r = `r round(coefs_dec_lms$estimate[coefs_dec_lms$yr==2015], 2)`, p = `r round(coefs_dec_lms$p.value[coefs_dec_lms$yr==2015], 2)`). 
On the other hand, in 2010, predicted sensitivities had no significant association with deciduousness (r = `r round(coefs_dec_lms$estimate[coefs_dec_lms$yr==2010], 2)`, p = `r round(coefs_dec_lms$p.value[coefs_dec_lms$yr==2010], 2)`). 
Moreover, models of all trees with TWI and deciduousness as predictors showed that the conditional effect of deciduousness on sensitivity (controlling for topographic wetness) was significant and negative in 2015 (`r round(isocline_coefs_dec$median[isocline_coefs_dec$yr==2015],2)`; 90% CI `r round(isocline_coefs_dec$lwr[isocline_coefs_dec$yr==2015],2)`, `r round(isocline_coefs_dec$upr[isocline_coefs_dec$yr==2015],2)`) but not in 2010 (`r round(isocline_coefs_dec$median[isocline_coefs_dec$yr==2010],2)`, 90% CI `r round(isocline_coefs_dec$lwr[isocline_coefs_dec$yr==2010],2)`, `r round(isocline_coefs_dec$upr[isocline_coefs_dec$yr==2010],2)`).

## Microenvironment and individual effects {-}

Individual tree responses to drought were not correlated with each other (Fig S6).

TWI had no significant on sensitivity in 2010 but a positive effect in 2015. 
Specifically, across all trees, the model with TWI and deciduousness showed that TWI had no effect on drought sensitivity in 2010 (median effect =`r round(isocline_coefs_twi$median[isocline_coefs_twi$yr==2010],2)`, 90% CI `r round(isocline_coefs_twi$lwr[isocline_coefs_twi$yr==2010],2)`, `r round(isocline_coefs_twi$upr[isocline_coefs_twi$yr==2010],2)`) but a positive effect in 2015 (median effect =`r round(isocline_coefs_twi$median[isocline_coefs_twi$yr==2015],2)`, 90% CI `r round(isocline_coefs_twi$lwr[isocline_coefs_twi$yr==2015],2)`, `r round(isocline_coefs_twi$upr[isocline_coefs_twi$yr==2015],2)`). 
The interaction between TWI and deciduousness was not significant in 2010 (median effect =`r round(isocline_coefs_inter$median[isocline_coefs_inter$yr==2010],2)`, 90% CI `r round(isocline_coefs_inter$lwr[isocline_coefs_inter$yr==2010],2)`, `r round(isocline_coefs_inter$upr[isocline_coefs_inter$yr==2010],2)`) but was significantly negative in 2015 (median effect =`r round(isocline_coefs_inter$median[isocline_coefs_inter$yr==2015],2)`, 90% CI `r round(isocline_coefs_inter$lwr[isocline_coefs_inter$yr==2015],2)`, `r round(isocline_coefs_inter$upr[isocline_coefs_inter$yr==2015],2)`).
Combined, predicted sensitivities of deciduous trees was more negative than evergreen trees in 2015, with this effect stronger in wetter sites (Fig `r fig_species`b).

The differing effects of TWI on sensitivity in the two droughts, along with the influence of deciduousness, were confirmed by models that controlled for crown exposure and DBH as well as species variation in intercept and slopes (Fig `r fig_microenvs`a). 
These models showed small negative sensitivities in wetter sites in 2010 (median effect =`r round(coefs_spre_twi$median[coefs_spre_twi$yr==2010],2)`, 90% CI `r round(coefs_spre_twi$lwr[coefs_spre_twi$yr==2010],2)`, `r round(coefs_spre_twi$upr[coefs_spre_twi$yr==2010],2)`), but a strong positive effect of TWI in 2015 (median effect =`r round(coefs_spre_twi$median[coefs_spre_twi$yr==2015],2)`, 90% CI `r round(coefs_spre_twi$lwr[coefs_spre_twi$yr==2015],2)`, `r round(coefs_spre_twi$upr[coefs_spre_twi$yr==2015],2)`). 
The positive effect of TWI in 2015 was, however, mainly due to the response of evergreen species; deciduousness had a significant negative correlation with the slopes of the TWI effect in 2015 (Fig `r fig_microenvs`b). 
Predictions across the range of observed variables also showed that deciduous species generally had larger negative sensitivities in 2015, which did not  increase with TWI (Fig `r fig_microenvs`c). 
On the other hand, neither the mean sensitivities across species nor their response to TWI were associated with deciduousness in 2010.

![***Figure `r fig_microenvs`: Drivers of variation in sensitivity among individuals***. All panels represent model results from models (one for each year) predicting sensitivity with microenvironment conditions with species random effect on all slopes. a) Coefficient plots showing median effects and 90% CI for wetness, exposure and DBH of tree. b) wetness effect across all species and its relationship with deciduousness value. Line represents a significant correlation. c) Predicted relationship of drought sensitivity with topographic wetness index across all species derived from model predictions. Black line represents overall relationship (panel a); coloured lines for each species reflects deciduousness values. For comments - https://github.com/forestgeo/growth-precip-thailand/issues/13](display/Fig4.png)

@@

DBH was a significant predictor of crown exposure, with larger DBH associated with higher exposure categories in 2010 (median effect =`r round(coefs_spre_ordcii$median[coefs_spre_ordcii$yr==2010],2)`, 90% CI `r round(coefs_spre_ordcii$lwr[coefs_spre_ordcii$yr==2010],2)`, `r round(coefs_spre_ordcii$upr[coefs_spre_ordcii$yr==2010],2)`) and 2015 (median effect =`r round(coefs_spre_ordcii$median[coefs_spre_ordcii$yr==2015],2)`, 90% CI `r round(coefs_spre_ordcii$lwr[coefs_spre_ordcii$yr==2015],2)`, `r round(coefs_spre_ordcii$upr[coefs_spre_ordcii$yr==2015],2)`) (Fig 5a). 
Controlling for this direct effect of exposure, the remaining effect of DBH *per se* was significantly positive in 2010, while it was weakly negative in 2015 (Fig `r fig_microenvs`a, Fig `r fig_dag`a).
